An Empirical Analysis of the Role of Energy in Economic Growth

### 2013-05-03 EconGrowth.Rnw
Edits on the implications section.

Figure \ref{fig:Historical_Lattice_Graph} indicates the importance of energy quantification for economic growth modeling. Thermal energy ($q$) and exergy ($x$) grow at similar rates, unless there is a significant raw energy carrier substitution in the economy (such as China's rapid increase of natural gas and coal consumption relative to petroleum since 2000). The growth rate of $u$ is significantly different from $q$ and $x$ for most countries, however. \citet{Warr:2012cg} show that from 1950-1990, useful work ($u$) grew much faster than $q$, $x$, $k$, or $l$ due to efficiency gains in the U.S. economy. The results of \citeauthor{Warr:2012cg} suggest that the time-dependent technological progress term ($A$) in the Cobb-Douglas and CES production functions (Equations \ref{eq:CD} -- \ref{eq:CESe}) may be a proxy for increasing energy efficiency in the U.S. economy.

@@

<<CDeGridLattice, fig.width=ninePanelGraphWidth, fig.height=ninePanelGraphHeight, fig.cap="Grid search over $\\gamma$ for Cobb-Douglas with energy model.">>=
cdeGridLattice(energyType="Q")
@

@@

@@

************* Do we need this? **************

************* Martin needs to write this. **************

@@

Funding for Caleb Reese and Lucas Timmer was generously supplied by the Jack and Lois Kuipers Applied Mathematics Endowment and the Calvin College Alumni Association. Support for Matthew Kuperus Heun was provided by a Calvin College Research Followship. Partial funding for Martin de Wit was provided by the South African National Research Foundation (NRF).\footnote{Any opinion, findings, and conclusions or recommendations expressed herein are those of the authors, and, therefore, NRF does not accept any liability in regard thereto.} The authors are grateful to Benjamin S. Warr (INSEAD) for providing the REXS database \citep{Warr:2012a}. The authors thank Loren L. Heun (Western Michigan University) for insightful critique of the statistical analyses presented herein.

@@

In the spirit of Reproducible Research \citep{Gandrud:2013vx}, all data, spreadsheets, \texttt{R} code \citep{R}, and other materials associated with this paper can be found at \protect\url{https://github.com/MatthewHeun/Econ-Growth- R-Analysis}.

>>>>

Figure \ref{fig:Historical_Lattice_Graph} indicates the importance of energy quantification for economic growth modeling. Thermal energy ($q$) and exergy ($x$) grow at similar rates, unless there is a significant raw energy carrier substitution in the economy (such as China's rapid increase of coal and natural gas consumption relative to petroleum since 2000). The growth rate of $u$ is significantly different from $q$ and $x$ for most countries, however. \citet{Warr:2012cg} show that from 1950-1990, useful work ($u$) grew much faster than $q$, $x$, $k$, or $l$ due to efficiency gains in the U.S. economy. The results of \citeauthor{Warr:2012cg} suggest that the time-dependent technological progress term ($A$) in the Cobb-Douglas and CES production functions (Equations \ref{eq:CD} -- \ref{eq:CESe}) may be a proxy for increasing energy efficiency in the U.S. economy.

@@

<<CDeGridLattice, fig.width=ninePanelGraphWidth, fig.height=ninePanelGraphHeight, fig.cap="Grid search over $\\gamma$ for Cobb-Douglas model with thermal energy ($q$).">>=
cdeGridLattice(energyType="Q")
@

@@

The above analysis points to several implications.

\begin{itemize}
  \item{The use of exergy, as opposed to thermal energy, in economic growth models is probably unnecessary. The exergy allocation factors are all close to unity (Table \ref{tab:Exergy_Allocation}). In the absence of very large changes in energy mix for an economy, indexed exergy ($x$) and thermal energy ($q$) are very similar. Figure \ref{fig:Historical_Lattice_Graph} shows that $q$ and $x$ lines lie atop one another for almost every economy.}
  \item{Figure \ref{fig:CDeGridLattice} shows that the set of parameters ($\lambda$, $\alpha$, $\beta$, and $\gamma$) that minimizes $SSE$ for the energy-augmented Cobb-Douglas production function always (with the exception of Saudi Arabia) has $\beta = 0$ or $\gamma = 0$. This result implies that it is unnecessary to include both labor ($l$) and energy ($e$) in the Cobb-Douglas production function.}
  \item{Partial residual plots for both the Cobb-Douglas model (Figures \ref{fig:CDPartialResidualsQ}--\ref{fig:CDPartialResidualsU} and CES model (Figures \ref{fig:CESPartialResidualsQ}--\ref{fig:CESPartialResidualsU}) indicate that augmenting the Cobb-Douglas and CES production functions with energy is not beneficial.}
  \item{Figure \ref{fig:CIvsParam_Lattice_Graph} shows that important energy parameters for both the energy-augmented Cobb-Douglas ($\gamma$) and CES ($\delta$) production functions have large confidence intervals such that we are unable to say (at the 95\% confidence level) that energy is an important factor for economic growth, vis-\`{a}-vis capital stock ($k$) or labor ($l$) for most economies.}
  \item{Figure \ref{fig:CIvsParam_Lattice_Graph} indicates that energy (in all quantifications, $q$, $x$, and $u$) is complementary to capital stock ($k$) and labor ($l$) for producing economic growth.}
  \item{Figures \ref{fig:LINEXq_Params_Graph}--\ref{fig:LINEXu_Params_Graph} indicate that LINEX model fitting parameters $a_0$ and $c_t$ are significant (at the 95\% confidence level) for nearly all countries. The LINEX production function provides reasonable good fits to historical data. This evidence indicates that the LINEX production function (with energy ($e$) as its only factor of production) is a reasonable replacement for capital stock ($k$) and labor ($l$) in the Cobb-Douglas and CES production functions.}
  \item{The concept of useful work ($u$) has potential and should be explored beyond developed economies. Figure \ref{fig:Historical_Lattice_Graph} shows that $u$ grows at very different rates from $k$ or $l$ for the developed economies. The covariance values in Section \ref{sec:Covariance_Tables} show that $u$ has much lower covariances with $k$ and $l$ than $q$ or $x$ has with $k$ and $l$. Thus, $u$ may contain more information relative to $k$ and $l$ than $q$ or $x$ and may, therefore, be more useful in economic growth modeling than $q$ or $x$. ******* Check this assumption after we obtain covariance values for $u$. }
\end{itemize}

The evidence above leads to a re-conceptualization of the relationship among energy, capital stock, and labor: energy is \emph{embedded} in capital stock and labor, and, therefore, it is unwarranted and unnecessary to include \emph{both} energy and the capital stock/labor pair in production functions. The empirical evidence indicates that an energy-only production function (LINEX) also provides a reasonable model for economic growth.

************* The above section needs to be re-evaluated after the statistical work is tightened up. ****************

************* Martin needs to flesh out the above section. **************

@@

@@

Funding for Caleb Reese and Lucas Timmer was provided by the Jack and Lois Kuipers Applied Mathematics Endowment and the Calvin College Alumni Association. Support for Matthew Kuperus Heun was provided by a Calvin College Research Followship. Partial funding for Martin de Wit was provided by the South African National Research Foundation (NRF).\footnote{Any opinion, findings, and conclusions or recommendations expressed herein are those of the authors, and, therefore, NRF does not accept any liability in regard thereto.} The authors are grateful to Benjamin S. Warr (INSEAD) for providing the REXS database \citep{Warr:2012a}. The authors thank Loren L. Heun (Western Michigan University) for insightful critique of the statistical analyses presented herein.

@@

In the spirit of Reproducible Research \citep{Gandrud:2013vx}, all data, spreadsheets, \texttt{R} code \citep{R}, and other materials associated with this paper can be found at \protect\url{https://github.com/MatthewHeun/Econ-Growth-R-Analysis}.
